---
title: "Pull Macosko et al. single cell mouse retina data"
output:
  html_notebook:
author: David McGaughey
date: 2017-09-05
---

David,
Please see the link below for the paper on Single-Cell retinal sequencing paper.  We can ask the following questions.
Whether Nlz2 present one or more clusters?
Whether Nlz2 present in Amacrine or ganglion cell clusters?
If present, what other transcriptional factors present in those clusters?

http://www.sciencedirect.com/science/article/pii/S0092867415005498

There is another paper on Single-cell sequencing done on Bipolar cell.  See the link below.
http://www.sciencedirect.com/science/article/pii/S0092867416310078

Thanks for your help.

Elan


https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63472
https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE63472&format=file&file=GSE63472%5FP14Retina%5FlogDGE%2Etxt%2Egz
https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE63472&format=file&file=GSE63472%5FP14Retina%5Fmerged%5Fdigital%5Fexpression%2Etxt%2Egz
http://mccarrolllab.com/wp-content/uploads/2015/05/retina_clusteridentities.txt

Three approaches tried:

1. Seurat
2. Monocle
3. Aaron Lun (https://bioconductor.org/help/workflows/simpleSingleCell/)

Seurat was the more performant option, compared to Monocle. I also had better success with clustering assignments and t-SNE. 

Seurat-based approach
```{r}
library(cowplot)
library(Seurat)
library(tidyverse)
dge <- read_tsv('~/Downloads//GSE63472_P14Retina_merged_digital_expression.txt')
dge_mat <- dge[,2:ncol(dge)] %>% as.matrix() %>%  Matrix(.,sparse = TRUE)
row.names(dge_mat) <- dge$gene
retina <- CreateSeuratObject(raw.data = dge_mat, min.cells = 0.001 * ncol(dge_mat), min.genes = 200)

mito.genes <- grep(pattern = "^MT-", x = rownames(x = retina@data), value = TRUE)
percent.mito <- colSums(as.matrix(retina@data[mito.genes, ])) / colSums(as.matrix(retina@data))
retina <- AddMetaData(object = retina, metadata = percent.mito, col.name = "percent.mito")
VlnPlot(object = retina, features.plot = c("nGene", "nUMI", "percent.mito"), nCol = 3)


# GenePlot is typically used to visualize gene-gene relationships, but can
# be used for anything calculated by the object, i.e. columns in
# object@data.info, PC scores etc.  Since there is a rare subset of cells
# with an outlier level of high mitochondrial percentage and also low UMI
# content, we filter these as well
par(mfrow = c(1, 2))
GenePlot(object = retina, gene1 = "nUMI", gene2 = "percent.mito")
GenePlot(object = retina, gene1 = "nUMI", gene2 = "nGene")

# We filter out cells that have unique gene counts over 6,000 or less than
# 900 Note that low.thresholds and high.thresholds are used to define a
# 'gate' -Inf and Inf should be used if you don't want a lower or upper
# threshold.

# this dramatically reduces the number of cells from ~45,000 to ~13,0000, but clustering is much better
# macosko et al clustered on ~13,000 initially, then did some fancy stuff to project the remaining 35,000 (mostly rod cells)
# onto the mapping
retina_superset <- FilterCells(object = retina, subset.names = c("nGene", "percent.mito"), 
                      low.thresholds = c(200, -Inf), high.thresholds = c(6000, 0.05))
retina <- FilterCells(object = retina, subset.names = c("nGene", "percent.mito"), 
                      low.thresholds = c(900, -Inf), high.thresholds = c(6000, 0.05))

# Normalize the data
retina <- NormalizeData(object = retina, normalization.method = "LogNormalize")

# Choose gene outliers on mean-variability plot
retina <- FindVariableGenes(object = retina, x.low.cutoff = 0, y.cutoff = 2)
length(x = retina@var.genes)


# Perform negative-binomial regression on the variable genes, this sets their value in retina@scale.data, which is used for PCA/clustering
# We only do this on the variable genes to save time, but you can do this genome-wide
# We treat mitochondrial percentage, batch, and nUMI as confounding variables,
# I am using the variable gene set calculated from the 'tight' FilterCell threshold, as these are a more stable set
retina <- ScaleData(object = retina, vars.to.regress = c("percent.mito", "nUMI","orig.ident"), genes.use = retina@var.genes, model.use = "negbinom")

# you can save the object at any time to save results, and can restore it back in using load()
save(retina, file = "./retina_seurat__1.Rdata")


# Run PCA with the IRLBA package (iteratively computes the top dimensions, dramatic increase in speed since we only use a fraction of the PCs anyways)
# if you see the warning "did not converge–results might be invalid!; try increasing maxit or fastpath=FALSE", try increasing maxit
retina <- RunPCA(object = retina, pc.genes = retina@var.genes, pcs.compute = 40, pcs.print = 1:10, maxit = 2000, weight.by.var = FALSE)

# elbow at 30
PCElbowPlot(object = retina, num.pc = 40)
PrintPCA(object = retina, pcs.print = 1:36)
PCHeatmap(object = retina, pc.use = 1:12,100)
#PCHeatmap(object = retina, pc.use = 13:24,100)
#PCHeatmap(object = retina, pc.use = 25:36,100)

VizPCA(object = retina, pcs.use = 1:2)

# save.SNN means that you can easily re-run with different resolution values.
retina <- FindClusters(object = retina, reduction.type = "pca", dims.use = 1:30, resolution = 1.2, save.SNN = TRUE)

# match PCs used to FindClusters above
retina <- RunTSNE(object = retina, dims.use = 1:30, do.fast = TRUE, perplexity=10, check_duplicates=F)
TSNEPlot(object = retina, do.label = TRUE)
```

Label clusters with Macosko et al. assignments
```{r}
# load Macosko cluster assignments for each cell
macosko_cluster = read_tsv('~/Downloads/retina_clusteridentities.txt', col_names = F) %>% data.frame()
row.names(macosko_cluster) <- macosko_cluster$X1
macosko_cluster <- macosko_cluster %>% select(X1, X2)
colnames(macosko_cluster) <- c('id','Macosko_Clusters')
macosko_cluster$Macosko_Clusters <- as.integer(macosko_cluster$Macosko_Clusters)
# https://github.com/olgabot/macosko2015/blob/master/notebooks/02_make_celltype_metadata.ipynb
naming <- data.frame(rbind(c(1, 'Horizontal cells'),
 c(2, 'Retinal ganglion cells'),
 c(3, 'Amacrine cells'),
 c(4, 'Amacrine cells'),
 c(5, 'Amacrine cells'),
 c(6, 'Amacrine cells'),
 c(7, 'Amacrine cells'),
 c(8, 'Amacrine cells'),
 c(9, 'Amacrine cells'),
 c(10, 'Amacrine cells'),
 c(11, 'Amacrine cells'),
 c(12, 'Amacrine cells'),
 c(13, 'Amacrine cells'),
 c(14, 'Amacrine cells'),
 c(15, 'Amacrine cells'),
 c(16, 'Amacrine cells'),
 c(17, 'Amacrine cells'),
 c(18, 'Amacrine cells'),
 c(19, 'Amacrine cells'),
 c(20, 'Amacrine cells'),
 c(21, 'Amacrine cells'),
 c(22, 'Amacrine cells'),
 c(23, 'Amacrine cells'),
 c(24, 'Rods'),
 c(25, 'Cones'),
 c(26, 'Bipolar cells'),
 c(27, 'Bipolar cells'),
 c(28, 'Bipolar cells'),
 c(29, 'Bipolar cells'),
 c(30, 'Bipolar cells'),
 c(31, 'Bipolar cells'),
 c(32, 'Bipolar cells'),
 c(33, 'Bipolar cells'),
 c(34, 'Muller glia'),
 c(35, 'Astrocytes'),
 c(36, 'Fibroblasts'),
 c(37, 'Vascular endothelium'),
 c(38, 'Pericytes'),
 c(39, 'Microglia')))
colnames(naming) <- c('Macosko_Clusters', 'Cell Type')
naming$Macosko_Clusters <- as.integer(naming$Macosko_Clusters)
macosko_cluster <- macosko_cluster %>% left_join(.,naming)
row.names(macosko_cluster) <- macosko_cluster$id
macosko_cluster <- macosko_cluster %>% select(Macosko_Clusters, `Cell Type`)
retina <- AddMetaData(object = retina, metadata = macosko_cluster, col.name = c('Macosko_Clusters', `Cell Type`))
retina <- AddMetaData(object = retina, metadata = macosko_cluster$`Cell Type`, col.name = 'Cell_Type')

#print with macosko clustering
TSNEPlot(object = retina, do.label = TRUE, group.by = 'Macosko_Clusters')


# save updated object
save(retina, file = "./retina_seurat__2.Rdata")
```

Plot marker genes and ZFP503.
```{r, fig.width=6, fig.height=4}
# plot individual genes
FeaturePlot(object = retina, features.plot = c("PAX6", "VSX2", "OPN1MW", "RLBP1","NRL","SLC17A6","CRX","ZFP503"), 
            cols.use = c("gray", 
                         "blue"))
```



Cells with ZFP503 Expression. 

`r table(retina@data['ZFP503',] > 0.1)['TRUE']` cells out of `r ncol(retina@data)` have detectable ZFP503 expression. 
```{r}
table(retina@data['ZFP503',] > 0.1)
```

Expression of key retinal markers in ZFP503 positive cells
```{r}
zfp503_pos_cells <- colnames(retina@data)[retina@data['ZFP503',] > 0.1]

retina@data[c("POU4F1", "CALB2", "ISL1", "LHX2","PAX6", "VSX2", "OPN1MW", "RLBP1","NRL","SLC17A6","CRX","ZFP503"), zfp503_pos_cells] %>% as.matrix() %>% data.frame() %>% rownames_to_column('Gene') %>% gather('cell_id', 'Expression', -Gene) %>% ggplot(aes(x=Gene, y=Expression)) + geom_violin() + geom_jitter(alpha=0.2) + theme(axis.text.x = element_text(angle = 90, hjust = 1))

retina@data[c("POU4F1", "CALB2", "ISL1", "LHX2","PAX6", "VSX2", "OPN1MW", "RLBP1","NRL","SLC17A6","CRX","ZFP503"), zfp503_pos_cells] %>% as.matrix() %>% data.frame() %>% rownames_to_column('Gene') %>% gather('cell_id', 'Expression', -Gene) %>% ggplot(aes(x=Gene, y=Expression)) + geom_boxplot() + geom_jitter(alpha=0.2) + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5))
```

Macosko Clustering cell type assignments for ZFP503

Over half are assigned as amacrine
```{r}
retina@meta.data[zfp503_pos_cells,'Cell Type'] %>% table()
```

Most expressed genes across the `r table(retina@data['ZFP503',] > 0.1)['TRUE']` cells
```{r}
retina@data[,zfp503_pos_cells] %>% 
  as.matrix() %>% 
  data.frame() %>% 
  rownames_to_column('Gene') %>% 
  gather('cell_id', 'Expression', -Gene) %>% 
  group_by(cell_id) %>% 
  top_n(10) %>% 
  pull(Gene) %>% 
  table() %>% 
  sort(decreasing = T)
```

Find differentially expressed genes in ZFP503 positive cells relative to non-ZFP503 expressing cells
```{r}
retina_zfp503 <- retina
zfp503_tf <- retina@data['ZFP503',] > 0.1
zfp503_tf <- gsub(TRUE, 1, zfp503_tf)
zfp503_tf <- gsub(FALSE, 0, zfp503_tf)
names(zfp503_tf) <- names(retina@ident)
zfp503_tf <- as.factor(zfp503_tf)
retina_zfp503@ident <- zfp503_tf

zfp503_markers <- FindMarkers(object = retina_zfp503, ident.1 = 1)

zfp503_markers %>% DT::datatable()

```


Impute categories for remaining ~30,000 cells not included in classification
```{r}
library(caret)
library(doMC)
registerDoMC(cores=8)
set.seed(1234)
trainIndex <- createDataPartition(retina@meta.data$res.1, p=0.5, times=1, list=T)
macosko_trainIndex <- trainIndex <- createDataPartition(retina@meta.data$Macosko_Clusters, p=0.5, times=1, list=T)
retinaTrain <- t(retina@scale.data[,trainIndex$Resample1])
retinaTest <- t(retina@scale.data[,-trainIndex$Resample1])
outcomesTrain <- as.factor(retina@meta.data$res.1[trainIndex$Resample1])
outcomesTest <- as.factor(retina@meta.data$res.1[-trainIndex$Resample1])
myControl <- trainControl(method = "repeatedcv", repeats=5, number = 10)

rf_mod <- train(x=retinaTrain, y=outcomesTrain, method = 'rf', trControl = myControl)
svm_mod <- train(x=retinaTrain, y=outcomesTrain, method = 'svmLinear', trControl = myControl)
LogitBoost_mod <- train(x=retinaTrain, y=outcomesTrain, method = 'LogitBoost', trControl = myControl)
bayes_mod <- train(x=retinaTrain, y=outcomesTrain, method = 'bayesglm', trControl = myControl)
#naive_bayes_mod <-  train(x=retinaTrain, y=outcomesTrain, method = 'naive_bayes', trControl = myControl)
```

svm works the best so far (still want to try keras neural net)
```{r}
pred <- predict(svm_mod, retinaTest)
postResample(pred=pred, obs=outcomesTest)
```